---
title: "BIOL116_Shiny-App-Script_Quantitative-Categorical"
date: "`r Sys.Date()`"
author: "Your Name Here"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = TRUE) # makes sure R code or messages are not printed
```

# Introduction & Instructions

This script is intended as an alternative to using the BIOL 116 Shiny App. As such, you can use it to produce figures, calculate descriptive statistics, and perform statistical analyses on your data. This specific script is intended to be used if your response (dependent) variable is **quantitative** and your explanatory (independent) variable is **categorical** 

In this `R` Markdown document the code used to perform actions such as producing figures is located within the `R` chunks. Alongside each line of code is a comment proceeded with a hashtag (#). The comments are descriptions of what the code is doing. **Things you are required to do are indicated with asterisks (*).**

Be sure to consult the rubric to ensure you have done everything that was asked for your assignment!

## Step 1: Set Your Working Directory

Follow the instructions [here](https://ubco-biology.github.io/Procedures-and-Guidelines/set-a-working-directory-in-rstudio.html) in the Procedures and Guidelines Document to set your working directory. 

## Step 2: Installing & Loading Required Packages

Copy the following code into the RStudio Console (lower left pane of RStudio) to install the necessary packages:

install.packages("ggplot2")
install.packages("dplyr")
install.packages("car")

Once the required packages are installed, you need to load these packages within your R Markdown document. To load the packages, run the `R` chunk below.  

```{r}

library("ggplot2")  
library("dplyr") 
library("car")

```

## Step 3: Uploading Your Data

Make sure your data is saved as a UTF-8 CSV file and is located within your working directory. In other words, your data should be saved in the same folder as the R Project you created in Step 1. 

* Replace `insert-data-file-name-here.csv` with the file name that your data is saved as to your computer

```{r}

my_data <- read.csv(file = "insert-data-file-name-here.csv", header = TRUE) 

# Reads your CSV file into R
# header = TRUE if the first row of values in your data has header information (aka column names)
# header = FALSE if there is no header information

```

## Step 4: Producing Figures

The two best options for displaying a categorical explanatory variable with a quantitative response variable are a boxplot and stripchart. Typically a stripchart is used when there are < 20 data points within each group, and a boxplot is used if there is > 20 data points in each group. 

* Replace **TWO** `insert-x-variable-name-here` with your explanatory (independent) variable's name

```{r}
my_data$insert-x-variable-name-here <- as.factor(my_data$insert-x-variable-name-here) # factoring the categorical variable

confint.fun.ttest <- function(x, conf = 0.95){ # defining a function to calculate the confidence interval
  return(data.frame(Mean = mean(x, na.rm = T),
                    ymin  = t.test(x, conf.level = conf)$conf.int[1],
                    ymax = t.test(x, conf.level = conf)$conf.int[2]))}
```

**Code for Creating a Boxplot**

* Replace `Insert your figure caption here` with your desired figure caption
* Replace **TWO** `insert-x-variable-name-here` with your explanatory (independent) variable's name
* Replace **TWO** `insert-y-variable-name-here` with your response (dependent) variable's name

```{r fig.cap="Insert your figure caption here."}

ggplot(my_data, 
    aes(x = insert-x-variable-name-here, y = insert-y-variable-name-here)) + # the variables to plot
    geom_boxplot(width= 0.5, fill = 'lightgrey') + # creates the boxplot, defines its size and colour
        stat_summary(fun.data = confint.fun.ttest, geom = 'errorbar', # creates the error bars for the confidence intervals, defines their color, size, and position
                     colour = 'black', width = 0.07, 
                     position = position_nudge(x = 0.4)) +
        stat_summary(fun = mean, # adds the mean to the graph as a point, defines colour, size, and position of the point
                     geom = 'point', 
                     colour = 'firebrick', 
                     size = 2, 
                     position = position_nudge(x = 0.4)) +
    labs(x = "insert-x-variable-name-here", y = "insert-y-variable-name-here") + # axes labels
    theme(
      axis.line = element_line(colour = 'gray', size = .75), # colour and size of axes
      panel.background = element_blank(), # no colour for plot area
      plot.background = element_blank()) # no colour for whole chart incl. behind axes

```

**Code for Creating a Stripchart**

* Replace `Insert your figure caption here` with your desired figure caption
* Replace **TWO** `insert-x-variable-name-here` with your explanatory (independent) variable's name
* Replace **TWO** `insert-y-variable-name-here` with your response (dependent) variable's name

```{r fig.cap="Insert your figure caption here."}
                    
ggplot(my_data,
       aes(x = insert-x-variable-name-here, y = insert-y-variable-name-here)) + # the variables to plot 
        geom_jitter(colour = 'black', size = 3, shape = 1, width = 0.1) + # creating data points, slightly shifting them to reduce overlap, defines colour and size of points
        stat_summary(fun.data = confint.fun.ttest, geom = 'errorbar', # creates the error bars for the confidence intervals, defines their color, size, and position
                     colour = 'black', width = 0.07, 
                     position = position_nudge(x = 0.15)) +
        stat_summary(fun = mean, # adds the mean to the graph as a point, defines colour, size, and position of the point
                     geom = 'point', 
                     colour = 'firebrick', 
                     size = 2, 
                     position = position_nudge(x = 0.15)) +
    labs(x = "insert-x-variable-name-here", y = "insert-y-variable-name-here") + # axes labels
    theme(
      axis.line = element_line(colour = 'gray', size = .75),
      panel.background = element_blank(),
      plot.background = element_blank())

```

**OPTIONAL:** 

If you are feeling ambitious, you are welcome to utilize [this resource](https://appsilon.com/ggplot2-boxplots/) for boxplots or this [one](http://www.sthda.com/english/wiki/ggplot2-scatterplot-easy-scatter-plot-using-ggplot2-and-r-statistical-software) for stripcharts to alter the script to further customize your figures. 

## Step 5: Calculating Descriptive Statistics

* Replace **ONE** `insert-x-variable-name-here` with your explanatory (independent) variable's name
* Replace **FOUR** `insert-y-variable-name-here` with your response (dependent) variable's name

```{r}
stats <- my_data %>%
          group_by(insert-x-variable-name-here) %>% # groups data by your categorical variable
        summarise(
          n = n(), # calculates sample size
          Mean = round(mean(insert-y-variable-name-here, na.rm = T), 1), # calculates the mean
          SD = round(sd(insert-y-variable-name-here, na.rm = T), 2), # calculates the standard deviation
          Median = round(median(insert-y-variable-name-here, na.rm = T), 1), # calculates the median
          iqr = round(IQR(insert-y-variable-name-here, na.rm = T), 2)) %>% # calculates the interquartile range
        as.data.frame() # creates a data frame of the descriptive stats    

stats # displays the data frame
```

## Step 6: Performing Statistical Analyses

The appropriate statistical test to assess for a relationship between a categorical and quantitative variable depends on how many groups the categorical variable has. If the categorical variable has exactly two groups the t-test is best. If the categorical variable has more than two groups then an ANOVA should be performed. 

* Once you have decided which test is best for you data, **DELETE** the R chunk for the opposite statistical test so it is not included in your final PDF output. 
    * For example, if your categorical variable has two groups and you want to perform a t-test: **KEEP** the R chunk that performs the t-test but **DELETE** the R chunk that performs the ANOVA 

**Code for t-Test**

* Replace **ONE** `insert-x-variable-name-here` with your explanatory (independent) variable's name
* Replace **ONE** `insert-y-variable-name-here` with your response (dependent) variable's name

```{r}
t.test(insert-y-variable-name-here ~ insert-x-variable-name-here, data = my_data, var.equal = TRUE, conf.level = 0.95)

# chooses the variables and data set, var.equal = TRUE indicates that the assumption of equal variance is met, defines the confidence interval used
```

**Code for ANOVA**

* Replace **ONE** `insert-x-variable-name-here` with your explanatory (independent) variable's name
* Replace **ONE** `insert-y-variable-name-here` with your response (dependent) variable's name

```{r}

lm <- lm(insert-y-variable-name-here ~ insert-x-variable-name-here, data = my_data) # creates a linear model using the variables from your data set
lm.anova <- anova(lm) # performs the ANOVA
lm.anova # displays the results 
```

## Step 7: Creating a PDF of the Results

Follow these steps to create a PDF of the results from your analysis.

* Click the dropdown arrow to the right of the Knit button
* Next click Knit to PDF
